---
title: "R Workshop Data Cleaning, Visualization, and Statistics"
author: "Jennifer Guthrie and Idowu Olawoye"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br> <br>

#### Load the necessary packages

```{r packages, echo=TRUE, warning=FALSE, message=FALSE}

#---- Load the necessary packages ----

# install.packages("epiDisplay")
library(epiDisplay)

# install.packages("dplyr")
library(dplyr)

#install.packages("ggplot2")
library(ggplot2)


# install.packages("car")  
library(car)             # Load the car package, which we'll use later to run Levene's test

```

<br>

#### Let's use the outbreak package

```{r data, echo=TRUE}

#---- Load food outbreak that is part of the epiDisplay package ----
data(Outbreak) 

# To understand the dataset view the information in the Help
#?Outbreak

# Display a summary of the dataset
summary(Outbreak)

```

<br> <br>

#### Case Definition

To identify which individuals were considered cases we need a Case Definition In this outbreak a case was defined as a person who had any of the four symptoms: 'nausea', 'vomiting', 'abdpain' or 'diarrhea'

```{r casedef, echo=TRUE}

# Let's create a variable that indicates who is a case and who is not
Outbreak <- Outbreak %>%
              mutate(case = ifelse(nausea == 1 |
                                     vomiting == 1 |
                                     abdpain == 1 | 
                                     diarrhea == 1, "Yes",
                                      "No"))
```

<br>

#### Case Counts

```{r casecounts, echo=TRUE}

# How many cases were there in the outbreak
Outbreak %>%
  group_by(case) %>%
  summarize(total = n()) %>%
  mutate(rel.freq = round(100*(total/sum(total)),1))

```

<br>

#### Case Order

```{r order, echo=TRUE}

# Notice that the order is No and then Yes (alphabetical), you can reorder
Outbreak <- Outbreak %>%
              mutate(case = factor(case, levels = c("Yes", "No")))


# Check that the reorder worked
Outbreak %>%
              group_by(case) %>%
              summarize(total = n()) %>%
              mutate(rel.freq = round(100*(total/sum(total)),1))

```

<br>

#### Was it the eclairs?

```{r eclair, echo=TRUE}

# Let's say we suspect that it was the eclair that gave people food poisoning
# Let's summarize the data for that variable
Outbreak %>%
  group_by(eclair) %>%
  summarize(total = n()) %>%
  mutate(rel.freq = round(100*(total/sum(total)),1))

```

<br>

#### How is the dataset coded?

What do the numbers mean under eclair? Are there people who ate 80 and 90 pieces of eclair!?

```{r eclaircoding, echo=TRUE}

# Perhaps we should summarize the data with the unknown info removed 
# from the denominator
Outbreak %>%
  filter(eclair != 80 & eclair != 90) %>%
  group_by(eclair) %>%
  summarize(total = n()) %>%
  mutate(rel.freq = round(100*(total/sum(total)),1))

```

<br>

#### Categorization

```{r eclaircats, echo=TRUE}

# Maybe we should also create some categories
Outbreak %>%
  filter(eclair != 80 & eclair != 90) %>%
  mutate(eclair.eaten = ifelse(eclair <= 0, "No",
                          ifelse(eclair > 0, "Yes",NA))) %>%
  mutate(eclair.eaten = factor(eclair.eaten, levels = c("Yes", "No"))) %>%
  group_by(eclair.eaten) %>%
  summarize(total = n()) %>%
  mutate(rel.freq = round(100*(total/sum(total)),1))

```

<br>

#### Analysis

```{r eclairanalysis, echo=TRUE, message=FALSE}

# How many of the cases ate the eclairs?
Outbreak %>%
  filter(eclair != 80 & eclair != 90) %>%
  mutate(eclair.eaten = ifelse(eclair <= 0, "No",
                          ifelse(eclair > 0, "Yes", NA))) %>%
  mutate(eclair.eaten = factor(eclair.eaten, levels = c("Yes", "No"))) %>%
  group_by(case, eclair.eaten) %>%
  summarize(total = n())

```

<br>

#### Data Visualization

##### Stacked barplot

```{r eclairviz, echo=TRUE, message=FALSE}

#---- Create a STACKED barplot to visualize the proportions ----
Outbreak %>%
  filter(eclair != 80 & eclair != 90) %>%
  mutate(eclair.eaten = ifelse(eclair <= 0, "No",
                               ifelse(eclair > 0, "Yes", NA))) %>%
  mutate(eclair.eaten = factor(eclair.eaten, levels = c("Yes", "No"))) %>%
  group_by(case, eclair.eaten) %>%
  summarize(total = n()) %>%
  mutate(rel.freq = round(100*(total/sum(total)),1)) %>%
    ggplot(aes(x = case, y = rel.freq, fill = eclair.eaten)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(x = "Case", y = "Total", title = "Cases by Eclair Eaten") +
    scale_fill_manual(values = c("Yes" = "#C934BD", "No" = "#34BDC9"), name = "Eclair Eaten") +
    theme_minimal()

```

<br>

#### Data Visualization

##### Side-by-Side barplot

```{r eclairviz2, echo=TRUE, message=FALSE}

#---- Create a SIDE-BY-SIDE barplot to visualize the proportions ----
Outbreak %>%
  filter(eclair != 80 & eclair != 90) %>%
  mutate(eclair.eaten = ifelse(eclair <= 0, "No",
                               ifelse(eclair > 0, "Yes", NA))) %>%
  mutate(eclair.eaten = factor(eclair.eaten, levels = c("Yes", "No"))) %>%
  group_by(case, eclair.eaten) %>%
  summarize(total = n()) %>%
  mutate(rel.freq = round(100*(total/sum(total)),1)) %>%
    ggplot(aes(x = case, y = rel.freq, fill = eclair.eaten)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(x = "Case", y = "Percentage (%)", title = "Cases by Eclair Eaten") +
    scale_fill_manual(values = c("Yes" = "#C934BD", "No" = "#34BDC9"), name = "Eclair Eaten") +
    theme_minimal()
    
```

<br>

#### Statistics

Is there a statistically higher number of people that ate eclairs that got sick?

```{r eclairvar, echo=TRUE, message=FALSE}

# Let's formally create the eclair.eaten variable
Outbreak.new1 <- Outbreak %>%
                  filter(eclair != 80 & eclair != 90) %>%
                  mutate(eclair.eaten = ifelse(eclair <= 0, "No",
                                        ifelse(eclair > 0, "Yes",NA))) %>%
                  mutate(eclair.eaten = factor(eclair.eaten, levels = c("Yes", "No")))



# Next we will make a contingency table
contingency_table <- table(Outbreak.new1$case, Outbreak.new1$eclair.eaten)

# Perform chi-squared test
chi_square_test <- chisq.test(contingency_table)

# Print results
print(chi_square_test)

    
```

<br>

#### Another example

```{r sexvar, echo=TRUE, message=FALSE}


# Summarise the cases by sex
Outbreak.new1 %>%
    group_by(sex) %>%
      summarize(total = n()) %>%
      mutate(rel.freq = round(100*(total/sum(total)),1))


# Go ahead create a new variable (sex.new) to change the coding of the 
# variable sex (0=F, 1=M) using the mutate function from the dplyr package
# create a new dataframe and name it Outbreak.new2
Outbreak.new2 <- Outbreak.new1 %>%
                  mutate(sex.new = ifelse(sex == 0, "F", "M"))

# Check that you recoded this new field correctly
Outbreak.new2 %>%
  group_by(sex, sex.new) %>%
  summarize(total = n())

```

<br>

#### Visualize

```{r boxplot, echo=TRUE, message=FALSE}

# Generate a boxplot to visualize age by sex
ggplot(Outbreak.new2, aes(x = sex.new, y = age)) +
  geom_boxplot() +
  labs(x = "Sex", y = "Age", title = "Boxplot of Age by Sex") +
  theme_minimal()
  
```

<br>

#### Statistics

```{r agesexstats, echo=TRUE, message=FALSE}

#--- Calculate the mean age of Males and Females
Outbreak.new2 %>%
  filter(age != 99) %>%
  group_by(sex.new) %>%
  summarize(mn = mean(age),
            sd = sd(age),
            md = median(age),
            Q1 = round(quantile(age, 0.25)), 
            Q3 = round(quantile(age, 0.75)))
  
```

<br>

#### Data Cleaning

```{r agefilter, echo=TRUE, message=FALSE}

# We should create a filtered dataset in which the records with missing age
# are removed
Outbreak.new3 <- Outbreak.new2 %>%
                  filter(age != 99)

```

<br>

#### Statistical Testing

```{r statsign, echo=TRUE, message=FALSE, warning=FALSE}

# Is there a statistically significant difference in age between males and females?

# Step 1: Check Assumptions
# Shapiro-Wilk test for normality
# If p ≤ 0.05: You reject the null hypothesis and conclude that the data are not normally distributed.
shapiro.test(Outbreak.new3$age[Outbreak.new3$sex.new == "M"])
shapiro.test(Outbreak.new3$age[Outbreak.new3$sex.new == "F"])

# Levene's test for homogeneity of variances
# p-value (≤ 0.05): Indicates evidence against the null hypothesis, 
# suggesting that the variances across groups are not equal.
leveneTest(age ~ sex.new, data = Outbreak.new2)


# Step 2: Perform Mann-Whitney U test 
# Non-parametric test because assumptions weren't met
mwu_test_result <- wilcox.test(age ~ sex.new, data = Outbreak.new3)

# Step 3: Print the test result
# p-value (≤ 0.05): Indicates that there is a significant difference 
# between males and females in their age distributions 
mwu_test_result


# If assumptions had been met the parametric test would be a t-test
# t_test_result <- t.test(age ~ sex.new, data = Outbreak.new2)
# t_test_result

```
